### 2022-03-02 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_networkary_classifiers@975948874c7ec87793df7c94132500727cfc4872 🚀
Deploying to gh-pages from @ PoisotLab/ms_networkary_classifiers@97286fcf1eae26f80ccdff7614dc8e698c1a6f82 🚀

Ecological networks are a backbone for key ecological and evolutionary
processes; yet enumerating all of the interactions they contain is a
daunting task, as it scales with \(S^2\), \emph{i.e.} the squared
species richness TODO. Recent contributions to the field of ecological
network prediction (Pichler et al., 2020; Strydom et al., 2021;
\textbf{Becker2021OptPre?}) highlight that although interactions can be
predicted by adding ecologically relevant information (in the form of,
\emph{e.g.} traits), we do not have robust guidelines as to how the
predictive ability of these models should be evaluated, nor about how
the models should be trained. Here, by relying on simple derivations and
a series of simulations, we formulate a number of such guidelines,
specifically for the case of binary classifiers derived from thresholded
values. Specifically, we conduct an investigation of the models in terms
of their skill (ability to make the right prediction), bias (trends
towards systematically over-predicting one class), class imbalance (the
relative number of cases representing interactions), and show how these
effects interact. We conclude on the fact that models with the best
interaction-scale predictive score do not necessarily result in the most
accurate representation of the network.

@@

In the following section, we will generate random bipartite networks
(this works without loss of generality on unipartite networks), and
train four binary classifiers (as well as an ensemble model using the
sum of ranged outputs from the component models) on 30\% of the
interaction data. In practice, testing usually uses 70\% of the total
data; for ecological networks, where interactions are sparse \emph{and}
the number of species is low, this may not be the best solution, as the
testing set becomes constrained not by the \emph{proportion} of
interactions, but by their \emph{number}. Preliminary experiments using
different splits revealed no qualitative change in the results. Networks
are generated by picking a random infectiousness trait \(v_i\) for 100
species (from a beta distribution \(B(\alpha=6,\beta=8)\) distribution),
and a resistance trait \(h_j\) for 100 species (from
\(B(\alpha=2,\beta=8)\) distribution). There is an interaction between
\(i\) and \(j\) when \(v_i-\xi/2 \le h_j \le v_i+\xi/2\), where \(\xi\)
is a constant regulating the connectance of the network (there is an
almost 1:1 relationship between \(\xi\) and connectance), and varies
uniformly in \([0.05, 0.35]\). This model gives fully interval networks
that are close analogues to the bacteria--phage model of Weitz et al.
(2005), with both a modular structure and a non-uniform degree
distribution. This dataset is easy for almost any algorithm to learn:
when trained with features \([v_i, h_j, \text{abs}(v_i, h_j)] ^T\) to
predict the interactions between \(i\) and \(j\), all four models
presented below were able to reach almost perfect predictions all the
time (data not presented here) -- this is in part because the rule
(there is maximum value of the distance between traits for which there
is an interaction) is fixed for all interactions. In order to make the
problem more difficult to solve, we use \([v_i, h_j]\) as a feature
vector (\emph{i.e.} the traits on which the models are trained), and
therefore the models will have to uncover that the rule for interaction
is \(\text{abs}(v_i, h_j) \le \xi\). The models therefore all have the
following form, where \(i_{i,j}\) is an interaction from species \(i\)
to species \(j\):

@@

The training sample is composed of 30\% of the \(10^4\) possible entries
in the network, \emph{i.e.} \(n=3000\). Out of these interactions, we
pick a proportion \(\nu\) (the training set balance) to be positive, so
that the training set has \(\nu n\) interactions, and \((1-\nu) n\)
non-interactions. We vary \(\nu\) uniformly in \(]0,1[\). This allows to
evaluate how the measures of binary classification performance respond
to artificially rebalanced dataset for a given network connectance. The
rest of the dataset (\(n=7000\) pairs of species) is used as a testing
set, on which all furher measures are calculated. Note that although the
training set is balanced, the testing set is not, and retains (part of)
the imbalance of the original data.

@@

\begin{figure}
\hypertarget{fig:optimbias}{%
\centering
\includegraphics{figures/optim_bias.png}
\caption{Value of the optimal training set balance for the different
models and measures evaluated here, over a range of connectances.
Informedness was reliably maximized for balanced training sets, and kept
this behavior across models. For other measures, larger connectances in
the true network allowed lower biases in the training set. In a large
number of cases, ``over-correcting'' by having training sets with more
than half instances representing interactions would maximize the values
of the model performance measures.}\label{fig:optimbias}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:optimvalue}{%
\centering
\includegraphics{figures/optim_value.png}
\caption{When trained on their optimally biased training set, most
models were able to maximize their performance; this is not true for
decision tree, which had a very low PR-AUC, and to some extent for ridge
regression who had a slow increase with network connectance. The
ensemble had a consistently high performance despite incorporating poor
models.}\label{fig:optimvalue}
}
\end{figure}

@@

Finally, it is noteworthy that the ensemble model was systematically
better than the component models; even when the models were individually
far form perfect, the ensemble was able to leverage the different biases
expressed by the models to make an overall more accurate prediction. We
do not expect that ensembles will \emph{always} be better than single
models. In a recent multi-model comparison, (\textbf{Becker2021OptPre?})
found that the ensemble was \emph{not} the best model. There is no
general conclusion to draw from this besides reinforcing the need to be
pragmatic about which models should be included in the ensemble, or
whether to use an ensemble at all. In a sense, the surprising peformance
of the ensemble model should form the basis of the last recommendation:
optimal training set bias and its interaction with connectance and
binary classifier is, in a sense, an hyperparameter that should be
assessed. The distribution of results in fig.~\ref{fig:optimbias} and
fig.~\ref{fig:optimvalue} show that there are variations around the
trend; furthermore, networks with different structures than the one we
simulated here may respond in different ways.

>>>>

Ecological networks are a backbone for key ecological and evolutionary
processes; yet enumerating all of the interactions between \(S\) species
is a daunting task, as it scales with \(S^2\), \emph{i.e.} the squared
species richness (\textbf{Martinez1992ConCon?}). Recent contributions to
the field of ecological network prediction (Pichler et al., 2020;
Strydom et al., 2021; \textbf{Becker2022OptPre?}) highlight that
although interactions can be predicted by adding ecologically relevant
information (in the form of, \emph{e.g.} traits), we do not have robust
guidelines as to how the predictive ability of these models should be
evaluated, nor about how the models should be trained. Here, by relying
on simple derivations and a series of simulations, we formulate a number
of such guidelines, specifically for the case of binary classifiers
derived from thresholded values. Specifically, we conduct an
investigation of the models in terms of their skill (ability to make the
right prediction), bias (trends towards systematically over-predicting
one class), class imbalance (the relative number of cases representing
interactions), and show how these effects interact. We conclude on the
fact that models with the best interaction-scale predictive score do not
necessarily result in the most accurate representation of the network.

@@

In the following section, we will generate random bipartite networks,
and train four binary classifiers (as well as an ensemble model using
the sum of ranged outputs from the component models) on 50\% of the
interaction data. In practice, testing usually uses 70\% of the total
data; for ecological networks, where interactions are sparse \emph{and}
the number of species is low, this may not be the best solution, as the
testing set becomes constrained not by the \emph{proportion} of
interactions, but by their \emph{number}. Preliminary experiments using
different splits revealed no qualitative change in the results. Networks
are generated by picking a random infectiousness trait \(v_i\) for 100
species (from a beta distribution \(B(\alpha=6,\beta=8)\) distribution),
and a resistance trait \(h_j\) for 100 species (from
\(B(\alpha=2,\beta=8)\) distribution). There is an interaction between
\(i\) and \(j\) when \(v_i-\xi/2 \le h_j \le v_i+\xi/2\), where \(\xi\)
is a constant regulating the connectance of the network (visual
exploration of the parameters show that there is an almost 1:1
relationship between \(\xi\) and connectance), and varies uniformly in
\([0.05, 0.35]\). This model gives fully interval networks that are
close analogues to the bacteria--phage model of Weitz et al. (2005),
with both a modular structure and a non-uniform degree distribution.
This dataset is easy for almost any algorithm to learn: when trained
with features \([v_i, h_j, \text{abs}(v_i, h_j)] ^T\) to predict the
interactions between \(i\) and \(j\), all four models presented below
were able to reach almost perfect predictions all the time (data not
presented here) -- this is in part because the rule (there is maximum
value of the distance between traits for which there is an interaction)
is fixed for all interactions, and any method able to learn non-linear
relationships should infer it without issues. In order to make the
problem more difficult to solve, we use \([v_i, h_j]\) as a feature
vector (\emph{i.e.} the traits on which the models are trained), and
therefore the models will have to uncover that the rule for interaction
is \(\text{abs}(v_i, h_j) \le \xi\). The models therefore all have the
following form, where \(i_{i,j}\) is an interaction from species \(i\)
to species \(j\):

@@

The training sample is composed of 50\% of the \(10^4\) possible entries
in the network, \emph{i.e.} \(n=5000\). Out of these interactions, we
pick a proportion \(\nu\) (the training set balance) to be positive, so
that the training set has \(\nu n\) interactions, and \((1-\nu) n\)
non-interactions. We vary \(\nu\) uniformly in \(]0,1[\). This allows to
evaluate how the measures of binary classification performance respond
to artificially rebalanced dataset for a given network connectance. The
rest of the dataset is used as a testing set, on which all further
measures are calculated. Note that although the training set is balanced
arbitrarily, the testing set is assembled so that it has the exact
connectance of the entire network; this ensures that the model is
evaluated under the class imbalance where the predictions will be made,
which represents a more meaningful evaluation. Note also that although
the simulated networks are bipartite, the algorithms have no
``knowledge'' of the network structure, and simply look at pairs of
species; therefore, the approach outlined here would also work for
unipartite networks.

@@

\begin{figure}
\hypertarget{fig:optimbias}{%
\centering
\includegraphics{figures/optimal_bias.png}
\caption{Value of the optimal training set balance for the different
models and measures evaluated here, over a range of connectances.
Informedness was reliably maximized for balanced training sets, and kept
this behavior across models. For other measures, larger connectances in
the true network allowed lower biases in the training set. In a large
number of cases, ``over-correcting'' by having training sets with more
than half instances representing interactions would maximize the values
of the model performance measures.}\label{fig:optimbias}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:optimvalue}{%
\centering
\includegraphics{figures/optimal_value.png}
\caption{When trained on their optimally biased training set, most
models were able to maximize their performance; this is not true for
decision tree, which had a very low PR-AUC, and to some extent for ridge
regression who had a slow increase with network connectance. The
ensemble had a consistently high performance despite incorporating poor
models.}\label{fig:optimvalue}
}
\end{figure}

@@

Finally, it is noteworthy that the ensemble model was systematically
better than the component models; even when the models were individually
far form perfect, the ensemble was able to leverage the different biases
expressed by the models to make an overall more accurate prediction. We
do not expect that ensembles will \emph{always} be better than single
models. In a recent multi-model comparison, (\textbf{Becker2022OptPre?})
found that the ensemble was \emph{not} the best model. There is no
general conclusion to draw from this besides reinforcing the need to be
pragmatic about which models should be included in the ensemble, or
whether to use an ensemble at all. In a sense, the surprising peformance
of the ensemble model should form the basis of the last recommendation:
optimal training set bias and its interaction with connectance and
binary classifier is, in a sense, an hyperparameter that should be
assessed. The distribution of results in fig.~\ref{fig:optimbias} and
fig.~\ref{fig:optimvalue} show that there are variations around the
trend; furthermore, networks with different structures than the one we
simulated here may respond in different ways.
